---
title: "README: NitrateExplorer"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NitrateExplorer

<!-- badges: start -->
<!-- badges: end -->

NitrateExplorer is an R package. It helps you explore NEON surface-water nitrate data. The package has a Shiny app and simple functions for summary and plotting. The data used in the app is included in the package.

## Installation

You can install the package from [GitHub](https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-Jing0922). Run this in R:

``` r
# install remotes if you do not have it
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")


# install the package from GitHub
remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-Jing0922")
```


You can install the development version of NitrateExplorer from [GitHub](https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-Jing0922) with:

``` r
# install.packages("pak")
pak::pak("ETC5523-2025/assignment-4-packages-and-shiny-apps-Jing0922")
```

## Website

A pkgdown website for this package is published at: [https://etc5523-2025.github.io/assignment-4-packages-and-shiny-apps-Jing0922/]( https://etc5523-2025.github.io/assignment-4-packages-and-shiny-apps-Jing0922/)



## Run the Shiny app

After install, run:

```{r}
library(NitrateExplorer)
```

```{r run app}
#| eval: false
run_nitrate_app()
```

This will open the interactive app. The app lets you choose sites, dates, and plot types. It also allows download of filtered data and plots.


## Main functions

The package includes these main functions:

- `run_app()`: launch the Shiny app.

- `calculate_daily_stats(data)` : return daily summaries.

- `calculate_monthly_stats(data)` : return monthly summaries.

- `calculate_yearly_stats(data)` : return yearly summaries.

- `plot_time_series(agg_data, sites, aggregation)` : draw time series plot from aggregated data.

- `boxplot_comparison(data, sites)` : draw boxplots to compare sites.

- `plot_distribution(data, sites, bin_size, show_density, facet_sites)` : draw distribution plot.

All functions are documented. See the `man/` folder for details and examples.


## Data

The package includes the `nitrate_clean` dataset containing quality-controlled nitrate measurements from NEON sites. It is saved with `usethis::use_data()` in `data/`. When using this data, please cite NEON appropriately according to their data use policies.

Key fields:

- siteID: ARIK, COMO, KING, LEWI, MAYF

- startDate: date of sample.

- surfWaterNitrateMean: average nitrate concentration (mg/L).

- finalQF: quality flag (0 = pass, 1 = fail).

The cleaning script is in data-raw/Nitrate-surface.R.


## Vignette and help

A vignette explains the main workflows. View the full vignette on our documentation website:
[**Nitrate Explorer Guide**](https://etc5523-2025.github.io/assignment-4-packages-and-shiny-apps-Jing0922/articles/nitrate_explorer_guide.html)


If you need more help, read the function help pages:
```{r}
#| eval: false
?calculate_monthly_stats
?plot_time_series
```


## Citation and source

Data source: NEON (National Ecological Observatory Network). Nitrate in surface water (DP1.20033.001), RELEASE-2025. https://doi.org/10.48443/wwa3-p420. Dataset accessed from https://data.neonscience.org/data-products/DP1.20033.001/RELEASE-2025 on October 27, 2025.

**Please cite NEON when you use the data.**


## Contributing

Contributions are welcome! Please feel free to submit pull requests or open issues for bugs and feature requests.

## License

MIT License (see file LICENSE).

## Contact

For questions or support, please open an issue on [GitHub](https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-Jing0922) or contact jwan0560\@student.monash.edu.
